---
title: "ps5-rmarkdown"
author: "Liuhan Ke"
date: "2023-02-15"
output: html_document
---

### Load And Check Data

#### 1. Load data
```{r}
library(dplyr)
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
```

#### 2. How many rows/columns do we have?

***The dataset contains `r nrow(gapminder)` rows and `r ncol(gapminder)` columns.***

#### 3. Print a small sample of data.
```{r}
gapminder %>% 
  sample_n(10)
```

### Descriptive Statistics

#### 1. How many countries are there in the dataset?
#### (Analyze all three: iso3, iso2 and name.)
```{r}
gapminder %>% 
  summarize(name_num = n_distinct(name),
            iso2_num = n_distinct(iso2),
            iso3_num = n_distinct(iso3))
```

#### 2. If you did this correctly, you saw that there are more names than iso-2 codes, and 
#### there are even more iso3 -codes. What is going on? Can you find it out?
##### a) Find how many names are there for each iso-2 code. Are there any iso-2 codes that 
##### correspond to more than one name? What are these countries?
```{r}
gapminder %>% 
  group_by(iso2) %>% 
  summarize(name_iso2_num = n_distinct(name)) %>% 
  arrange(desc(name_iso2_num))
```

***There are 2 NA values, which means no country appears more than once, but NA value***
***appears more than once.***

##### b) Now repeat the same for name and iso3-code. Are there country names that have more 
##### than one iso3-code? What are these countries?
##### Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}
gapminder %>% 
  group_by(name) %>% 
  summarize(name_iso3_num = n_distinct(iso3)) %>% 
  arrange(desc(name_iso3_num))
```

***There are 4 NA values in the country column, which means no country has more than one***
***iso3-code, or we cannot know which countries have that.***

#### 3. What is the minimum and maximum year in these data?
```{r}
gapminder %>% 
  filter(!is.na(time)) %>% 
  summarise(max_year = max(time),
            min_year = min(time))
```

### CO2 emissions

#### 1. How many missing co2 emissions are there for each year? Analyze both missing CO2 and 
#### co2_PC. Which years have most missing data?
```{r}
missing_co2 <- gapminder %>% 
  group_by(time) %>% 
  summarize(missing_num_co2 = sum(is.na(co2)),
            missing_num_co2_PC = sum(is.na(co2_PC)))
missing_co2
```

```{r}
missing_co2 %>% 
  mutate(rank = (dense_rank(desc(missing_num_co2)))) %>% 
  filter(rank == 1)
```

#### 2. Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
#### more countries of your choice. Explain what do you see.
```{r}
plot_emission <- gapminder %>%
  select(name, time, co2) %>% 
  filter(name %in% c("United States of America", "China", "India", 
                     "Botswana", "Canada", "Germany"), 
         !is.na(co2))
ggplot(plot_emission, aes(x = time, y = co2, col = name)) +
  geom_line() +
  labs(x = "Time (years)", y = "CO2 Emission (kt)",
       title = "Different Countries' CO2 Emission (kt) v.s. Time")
```

***Explanation: Before around 2005, the U.S. has the highest co2 emission, but after ***
***that, China has an sharp increase in CO2 emission and became the highest one. Botswana***
***has the lowest CO2 emission. Germany has the most missing number of CO2 emission.*** 
***(Because Germany only has about half of a line compared with other countries in the***
***graph)***

#### 3. Now let's analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the 
#### same countries. What does this figure suggest?
```{r}
plot_emission_PC <- gapminder %>%
  select(name, time, co2_PC) %>% 
  filter(name %in% c("United States of America", "China", "India", 
                     "Botswana", "Canada", "Germany"), 
         !is.na(co2_PC))
ggplot(plot_emission_PC, aes(x = time, y = co2_PC, col = name)) +
  geom_line() +
  labs(x = "Time (years)", y = "CO2 Emission (metric tons per capita)",
       title = "Different Countries' CO2 Emission (metric tons per capita) v.s. Time")
```

***Explanation: In this graph, I saw the U.S. has the highest emission, and Canada has*** 
***the second highest CO2 emission among the six countries. India has the lowest emission*** 
***overall, but Botswana was lower than India before 1970s. China has an sudden sharp*** 
***increase around 2005. Germany has the most missing number of CO2 emissions. (Because*** 
***Germany only has about half of a line compared with other countries in the graph)***

#### 4. Compute average CO2 emissions per capita across the continents (assume region is the 
#### same as continent). Comment what do you see.
##### Note: just compute averages over countries and ignore the fact that countries are of 
##### different size.
##### Hint: Americas 2016 should be 4.80.
```{r}
gapminder %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC), time == 2016) %>% 
  summarize(avg_co2_PC = mean(co2_PC))
```

#### 5. Make a barplot where you show the previous results--average CO2 emissions per capita 
#### across continents in 1960 and 2016.
```{r}
avg_co2_PC <- gapminder %>% 
  filter(!is.na(co2_PC), time %in% c(1960, 2016)) %>% 
  group_by(time, region) %>% 
  summarize(avg_co2_PC = mean(co2_PC))
ggplot(avg_co2_PC, aes(x = region, y = avg_co2_PC, fill = factor(time))) +
  geom_col(position = "dodge2") +
  labs(x = "Continent", y = "Average CO2 Emission (metric tons per capita)",
       fill = "Time", title = "Average CO2 Emission of 1960 and 2016 in Different Countries")
```

#### 6. Which countries are the three largest, and three smallest CO2 emitters (in terms of 
#### CO2 per capita) in 2016 for each continent? (Assume region is continent).
***(Because 2019 data is missing, I will use 2016)***
```{r}
gapminder %>% 
  filter(time == 2016, !is.na(region), !is.na(name), !is.na(co2_PC)) %>% 
  group_by(region) %>% 
  filter(dense_rank(desc(co2_PC)) <= 3 |
           dense_rank(co2_PC) <= 3) %>% 
  select(region, name, co2_PC) %>% 
  arrange(desc(co2_PC), .by_group = TRUE)
```

### GDP per capita

#### 1. Make a scatterplot of GDP per capita versus life expectancy by country, using data 
#### for 1960. Make the point size dependent on the country size, and color those according 
#### to the continent. Feel free to adjust the plot in other ways to make it better.
#### Comment what do you see there.
```{r}
gapminder %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), !is.na(region), time == 1960) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, col = region)) +
  geom_point(aes(size = totalPopulation)) +
  labs(title = "Life Expectancy v.s. GDP per capita for Different Regions",
       x = "GDP per capita", y = "Life Expectancy (year)")
```

***Comment: I found that as the GDP per capita increases, the life expectancy shows a*** 
***logarithmic growth. In general, Europe and Americas have the most GDP per capita and*** 
***longest life expectancy, while Africa and Asia have the least of them. In terms of*** 
***population, Asia has the most population.***

#### 2. Make a similar plot, but this time use 2019 data only.
```{r}
gapminder %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), !is.na(region), time == 2019) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, col = region)) +
  geom_point(aes(size = totalPopulation)) +
  labs(title = "Life Expectancy v.s. GDP per capita for Different Regions",
       x = "GDP per capita", y = "Life Expectancy (year)")
```

#### 3. Compare these two plots and comment what do you see. How has world developed through 
#### the last 60 years?

***In the two graphs, the GDP per capita and life expectancy both display a logarithmic*** 
***growth. However, the main region of the most GDP per capita and longest life expectancy***
***in 2016 changes from Europe and Americas to Europe, and the least changes from Asia and***
***Africa to Africa. In the last 60 years, Europe still thrives, Americas regress a little***
***bit, and Asia developed a lot.***

#### 4. Compute the average life expectancy for each continent in 1960 and 2019. Do the 
#### results fit with what do you see on the figures?
##### Note: here as average I mean just average over countries, ignore the fact that 
##### countries are of different size.
```{r}
gapminder %>% 
  filter(time %in% c(1960, 2019), !is.na(lifeExpectancy), !is.na(region)) %>% 
  group_by(time, region) %>% 
  summarize(avg_le = mean(lifeExpectancy))
```

***Yes, the result fits what I see in the figure.***

#### 5. Compute the average LE growth from 1960-2019 across the continents. Show the results 
#### in the order of growth. Explain what do you see.
##### Hint: these data (data in long form) is not the simplest to compute growth. But you may
##### want to check out the lag() function. And do not forget to group data by continent when
##### using lag(), otherwise your results will be messed up!

***I understand the 1960-2019 in the instruction as "from 1960 to 2019", so I did for*** 
***each year. ***
```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(region)) %>% 
  group_by(region, time) %>% 
  summarize(avg_le = mean(lifeExpectancy)) %>% 
  select(region, time, avg_le) %>%
  mutate(growth = avg_le - lag(avg_le)) %>% 
  arrange(desc(growth))
```

***Explanation: The life expectancy of Asia and Africa is increasing rapidly these*** 
***years.***

#### 6. Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both 
#### histograms on the same graph, see how well you can do it!
```{r}
gapminder %>% 
  filter(!is.na(GDP_PC), time %in% c(1960, 2019)) %>% 
  group_by(time) %>% 
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(position = "dodge2", bins = 20) +
  labs(x = "GDP per capita", y = "Frequency", 
       title = "GDP per capita for years of 1960 and 2019")
  
```

#### 7. What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When 
#### counting from top.)
```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name),
         time %in% c(1960, 2019)) %>% 
  group_by(time) %>% 
  mutate(rank = dense_rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  select(name, time, rank) %>% 
  arrange(desc(rank))
```

#### 8. If you did this correctly, then you noticed that US ranking has been falling quite a 
#### bit. But we also have more countries in 2019–what about the relative rank divided by the
#### corresponding number of countries that have LE data in the corresponding year?
```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name),
         time %in% c(1960, 2019)) %>% 
  group_by(time) %>% 
  mutate(rank = dense_rank(desc(lifeExpectancy)) / n_distinct(name)) %>% 
  filter(name == "United States of America") %>% 
  select(name, time, rank) %>% 
  arrange(desc(rank))
```
